---
title: "BiNet Preprocessing Tutorial"
author: "Monica Shen"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-overflow: wrap
    embed-resources: true
  md:
    toc: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

The *Bilingual Interactional Network Survey* (BiNet) extends traditional language background questionnaires (LHQs) by shifting the unit of analysis from the **individual** to the **relationship**. BiNet retains the individual-level measures while adding a relational layer: participants nominate the 15 people they interact with regularly, the language practices that characterize each relationship, and the ties among those nominees. This yields a multi-level dataset that can support both standard individual-differences analyses and network-based measures of bilingual language experience.

This tutorial script shows how to transform raw BiNet data exported from Network Canvas into analysis-ready datasets using R. The workflow is organized into the following stages:

1.  Load and merge Network Canvas export files with Language History Questionnaire (LHQ) data (`demo_lhq.csv`)
2.  Preprocess alter-level variables (recode, classify, and tidy)
3.  Build the `egor` object for egocentric network analysis
4.  Derive compositional measures describing network composition
5.  Save outputs for use in downstream analyses

Network visualization is covered in the companion script `../BiNet_Network_Visualization/BiNet_network_visualization.qmd`, which loads the output files saved in Step 5.

> **Note on demo data.** The Network Canvas export files and LHQ data used in this tutorial are fully simulated datasets created for instructional purposes. They do not contain real participant information.

> **Note on column names.** Column names in `demo_lhq.csv` will vary across studies depending on questionnaire design and data-collection platform. Customize the import and merge steps so that the participant identifier in your questionnaire (e.g., `participant_id`) matches the `ego_id` field in the Network Canvas export.

## **Core Concepts**

-   **ego:** the participant

-   **alters:** the people nominated by the participant

-   **ties:** the relationships among those alters

# Understanding the Data Structure

This script processes data exported from the **BiNet** Network Canvas protocol (`NetworkCanvasProtocol_BiNet_20260417`). Network Canvas exports **one set of three CSV files per participant**, all placed in a single folder (**Network Canvas Export**). All files for all participants are placed in a single folder. The script reads all matching files at once using `list.files()` + `lapply()` + `bind_rows()`.

### Three Levels

1.  [Ego-level Data](#ego): {ego_id}\_{networkCanvasSessionID}`_ego.csv`
2.  [Alter-level Data](#alter): {ego_id}\_{networkCanvasSessionID}`_attributeList_People.csv`
3.  [Tie-level Data](#ties): {ego_id}\_{networkCanvasSessionID}`_edgeList_tie.csv`

### Ego-level Data — `*_ego.csv` {#ego}

One file per participant; one row per file.

+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| Variable                    | Type       | Description                                                                                      |
+=============================+============+==================================================================================================+
| `networkCanvasEgoUUID`      | character  | Primary link variable that links `*_ego.csv`, `*_attributeList_People.csv`, `*_edgeList_tie.csv` |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| `networkCanvasCaseID`       | character  | case ID (same as `ego_id`)                                                                       |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| `networkCanvasSessionID`    | character  | Session identifier                                                                               |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| `networkCanvasProtocolName` | character  | Protocol filename (`NetworkCanvasProtocol_BiNet_20260417`)                                       |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| `sessionStart`              | datetime   | Session start timestamp (ISO 8601)                                                               |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| `sessionFinish`             | datetime   | Session end timestamp                                                                            |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| `sessionExported`           | datetime   | Export timestamp                                                                                 |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+
| `ego_id`                    | character  | 8-character participant id (e.g., `annbea01`) that links `*_ego.csv` and `lhq.csv`               |
+-----------------------------+------------+--------------------------------------------------------------------------------------------------+

**Notes**: `networkCanvasCaseID` and `ego_id` columns should be identical

### Alter-level Data — `*_attributeList_People.csv` {#alter}

One file per participant; 15 rows per file.

#### System columns

+------------------------+-----------+------------------------------------------------------------------------------------------+
| Variable               | Type      | Description                                                                              |
+========================+===========+==========================================================================================+
| `nodeID`               | integer   | Within-participant alter index (1–15)                                                    |
+------------------------+-----------+------------------------------------------------------------------------------------------+
| `networkCanvasEgoUUID` | character | Primary link variable that joins `*_attributeList_People.csv` back to `*_ego.csv`        |
+------------------------+-----------+------------------------------------------------------------------------------------------+
| `networkCanvasUUID`    | character | Unique alter identifier that links `*_attributeList_People.csv` and `*_edgeList_tie.csv` |
+------------------------+-----------+------------------------------------------------------------------------------------------+

#### Alter-level attributes

1.  **Alter demographics**

+----------------+-----------+----------------------------------------------------------------------+-------------------+
| Variable       | Type      | Values                                                               | Description       |
+================+===========+======================================================================+===================+
| `name`         | character | Free text                                                            | Alter name        |
+----------------+-----------+----------------------------------------------------------------------+-------------------+
| `alter_gender` | character | `female`, `male`, `other`                                            | Alter's gender    |
+----------------+-----------+----------------------------------------------------------------------+-------------------+
| `alter_age`    | character | `0_9`, `10_17`, `18_25`, `26_35`, `36_45`, `46_55`, `56_65`, `65_99` | Alter's age group |
+----------------+-----------+----------------------------------------------------------------------+-------------------+

2.  **Ego-alter Interactional attributes**

+-----------------------------------+----------+--------------+-----------------------------------------------------------------------------------------------------+
| Variable                          | Type     | Values       | Description                                                                                         |
+===================================+==========+==============+=====================================================================================================+
| `ego_alter_interaction_frequency` | ordinal  | 1–5          | 1 = at least once a year; 5 = at least once a day                                                   |
+-----------------------------------+----------+--------------+-----------------------------------------------------------------------------------------------------+
| `ego_alter_emotional_closeness`   | ordinal  | 1–5          | 1 = not close at all; 5 = extremely close                                                           |
+-----------------------------------+----------+--------------+-----------------------------------------------------------------------------------------------------+
| `ego_alter_cs_frequency`          | ordinal  | 1–4, or `""` | 1 = rarely; 4 = always.                                                                             |
|                                   |          |              |                                                                                                     |
|                                   |          |              | **Empty string when ego uses only one language with this alter** (stage skipped by protocol filter) |
+-----------------------------------+----------+--------------+-----------------------------------------------------------------------------------------------------+

**Interaction context** (dummy-coded, 5 columns)

**Format:** `ego_alter_interaction_context_{context}` \| **Type:** logical (`true`/`false`) \| **Note:** Exactly one column is `true` per alter (single-select CategoricalBin)

+-------------------------------------------+-------------------------------------------------------------+
| Column                                    | Context                                                     |
+===========================================+=============================================================+
| `ego_alter_interaction_context_family`    | Family (people you live with or close relatives)            |
+-------------------------------------------+-------------------------------------------------------------+
| `ego_alter_interaction_context_community` | Non-relatives sharing living spaces (housemates, neighbors) |
+-------------------------------------------+-------------------------------------------------------------+
| `ego_alter_interaction_context_school`    | Classmates, professors                                      |
+-------------------------------------------+-------------------------------------------------------------+
| `ego_alter_interaction_context_work`      | Coworkers, supervisors, clients                             |
+-------------------------------------------+-------------------------------------------------------------+
| `ego_alter_interaction_context_social`    | Recreational / hobby contexts outside school or work        |
+-------------------------------------------+-------------------------------------------------------------+

3\. Alter language repertoire (10 columns)

-   Format: `alter_language_{language}` \| Type: logical (`true`/`false`)

4\. Ego-alter language use (10 columns)

-   Format: `ego_alter_language_{language}` \| Type: logical (`true`/`false`)

Note: Language options included in the protocol: `english`, `spanish`, `mandarin`, `cantonese`, `filipino`, `vietnamese`, `arabic`, `korean`, `japanese`, `other`

### Tie-level data — `*_edgeList_tie.csv` {#ties}

One file per participant; variable number of rows per file.

+---------------------------+------------+------------------------------------------------------------------------------------------+
| Variable                  | Type       | Description                                                                              |
+===========================+============+==========================================================================================+
| `edgeID`                  | integer    | Within-participant edge index                                                            |
+---------------------------+------------+------------------------------------------------------------------------------------------+
| `from`                    | integer    | `nodeID` of source alter                                                                 |
+---------------------------+------------+------------------------------------------------------------------------------------------+
| `to`                      | integer    | `nodeID` of target alter                                                                 |
+---------------------------+------------+------------------------------------------------------------------------------------------+
| `networkCanvasEgoUUID`    | character  | Primary link variable that joins `*_edgeList_tie.csv` back to `*_ego.csv`                |
+---------------------------+------------+------------------------------------------------------------------------------------------+
| `networkCanvasUUID`       | character  | Unique alter identifier that links `*_attributeList_People.csv` and `*_edgeList_tie.csv` |
+---------------------------+------------+------------------------------------------------------------------------------------------+
| `networkCanvasSourceUUID` | character  | `networkCanvasUUID` of source alter                                                      |
+---------------------------+------------+------------------------------------------------------------------------------------------+
| `networkCanvasTargetUUID` | character  | `networkCanvasUUID` of target alter                                                      |
+---------------------------+------------+------------------------------------------------------------------------------------------+

**Notes:** `edgeID`, `from`, and `to` are unique only within a participant. Ties are undirected; source/target assignment is arbitrary.

## Linking the Pieces: Connect Multilevel Network Data

A key takeaway from the data structure above is that BiNet data are stored across **multiple levels** (ego, alter, and tie), but these files are not independent. They are connected through a small set of **linking variables** that allow researchers to reconstruct each participant’s personal network and merge information across files.

-   **`ego_id`**: participant ID (e.g., `annbea01`) used to link BiNet data with external datasets such as language history questionnaires, behavioral tasks, or demographic records

-   **`networkCanvasEgoUUID`**: identifies each participant (**ego**) within the Network Canvas system and links all three exported files

-   `networkCanvasUUID`: identifies each nominated person (**alter**) within an ego's network

-   `networkCanvasSourceUUID` / `networkCanvasTargetUUID`: define tie endpoints to specify the two alters involved in each tie

In short, although BiNet exports multiple files, the linking variables ensure they function as **one coherent relational dataset**.

## References

Network Canvas Documentation: https://documentation.networkcanvas.com/

`egor` package: https://cran.r-project.org/package=egor

## 0.Setup

```{r setup}
packages <- c("dplyr", "tidyr", "stringr", "readr", "egor",
              "janitor", "scales", "purrr")

installed_packages <- packages %in% rownames(installed.packages())
if (any(!installed_packages)) {
  install.packages(packages[!installed_packages])
}

invisible(lapply(packages, library, character.only = TRUE))
```

# Demo Script

## 1. Load and Merge Raw Data

Set `data_dir` to the folder containing all participant Network Canvas export files (`Network Canvas Export`). The `list.files()` → `lapply()` → `bind_rows()` pipeline reads all files matching each pattern and stacks them into a single data frame.

```{r load_data, results='hide'}
# Set to the folder containing all Network Canvas export files
data_dir <- "./Network Canvas Export" 

data_alter <- data_dir |>
  list.files(full.names = TRUE, pattern = "attributeList_People\\.csv$") |>
  lapply(readr::read_csv, show_col_types = FALSE) |>
  dplyr::bind_rows()

data_ego <- data_dir |>
  list.files(full.names = TRUE, pattern = "_ego\\.csv$") |>
  lapply(readr::read_csv, show_col_types = FALSE) |>
  dplyr::bind_rows()

data_edgelist <- data_dir |>
  list.files(full.names = TRUE, pattern = "edgeList_tie\\.csv$") |>
  lapply(readr::read_csv, show_col_types = FALSE) |>
  dplyr::bind_rows()

cat("Loaded", nrow(data_ego), "egos,",
    nrow(data_alter), "alters,",
    nrow(data_edgelist), "edges\n")
```

Before moving on, it is good practice to verify that the raw Network Canvas exports contain all the columns listed in the data dictionary above. The code below checks each of the three data frames (`data_ego`, `data_alter`, `data_edgelist`) against the expected column names. This check is designed to help ensure data quality and to make it easier to use this package and the accompanying script. **For the code to run smoothly end-to-end, the column names in your exports should match the ones specified in the data dictionary**. If your protocol uses different column names (e.g., a different set of focal languages, or added/removed attributes), update the `expected_*_cols` vectors below accordingly before running the rest of the script.

```{r sanity_check_columns}
# Expected columns per the BiNet data dictionary
expected_ego_cols <- c(
  "networkCanvasEgoUUID", "networkCanvasCaseID", "networkCanvasSessionID",
  "networkCanvasProtocolName", "sessionStart", "sessionFinish",
  "sessionExported", "ego_id"
)

expected_alter_cols <- c(
  # System columns
  "nodeID", "networkCanvasEgoUUID", "networkCanvasUUID",
  # Alter demographics
  "name", "alter_gender", "alter_age",
  # Ego–alter interactional attributes
  "ego_alter_interaction_frequency", "ego_alter_emotional_closeness",
  "ego_alter_cs_frequency",
  # Interaction context (5 dummy columns)
  "ego_alter_interaction_context_family",
  "ego_alter_interaction_context_community",
  "ego_alter_interaction_context_school",
  "ego_alter_interaction_context_work",
  "ego_alter_interaction_context_social",
  # Alter language repertoire (10 columns)
  paste0("alter_language_",
         c("english","spanish","mandarin","cantonese","filipino",
           "vietnamese","arabic","korean","japanese","other")),
  # Ego–alter language use (10 columns)
  paste0("ego_alter_language_",
         c("english","spanish","mandarin","cantonese","filipino",
           "vietnamese","arabic","korean","japanese","other"))
)

expected_edge_cols <- c(
  "edgeID", "from", "to",
  "networkCanvasEgoUUID", "networkCanvasUUID",
  "networkCanvasSourceUUID", "networkCanvasTargetUUID"
)

check_columns <- function(df, expected, label) {
  missing_cols <- setdiff(expected, names(df))
  extra_cols   <- setdiff(names(df), expected)

  cat("---", label, "---\n")
  cat("Rows:", nrow(df), "| Columns:", ncol(df), "\n")

  if (length(missing_cols) == 0) {
    cat("[OK] All expected columns present.\n")
  } else {
    cat("[MISSING] ", length(missing_cols), " column(s):\n", sep = "")
    cat(paste(" -", missing_cols), sep = "\n"); cat("\n")
  }

  if (length(extra_cols) > 0) {
    cat("[INFO] ", length(extra_cols), " extra column(s) not in spec:\n", sep = "")
    cat(paste(" -", extra_cols), sep = "\n"); cat("\n")
  }
  cat("\n")
  invisible(list(missing = missing_cols, extra = extra_cols))
}

check_columns(data_ego,      expected_ego_cols,   "Ego-level data (*_ego.csv)")
check_columns(data_alter,    expected_alter_cols, "Alter-level data (*_attributeList_People.csv)")
check_columns(data_edgelist, expected_edge_cols,  "Tie-level data (*_edgeList_tie.csv)")

stopifnot(
  all(expected_ego_cols   %in% names(data_ego)),
  all(expected_alter_cols %in% names(data_alter)),
  all(expected_edge_cols  %in% names(data_edgelist))
)
```

### 1.1 Load LHQ Data

The LHQ file contains ego-level self-report variables (demographics, language background) collected separately from Network Canvas. It is linked to the network data via `ego_id`.

```{r load_lhq}
# Set to the folder containing the LHQ file
lhq_dir <- "./"

lhq_df <- readr::read_csv(
  file.path(lhq_dir, "demo_lhq.csv"),
  show_col_types = FALSE
) |>
  dplyr::mutate(
    # Parse the comma-separated language string into a list column.
    # Used later when computing language homophily (§5.6).
    ego_lang_list = stringr::str_split(tolower(ego_lang), ",\\s*")
  )

dplyr::glimpse(lhq_df)
```

### 1.2 Build Linked Datasets

Three linked datasets are constructed here and used throughout the script:

-   **`egoData_linked`**: one row per participant; network session variables joined to LHQ variables via `ego_id`
-   **`alterData_linked`**: one row per alter; restricted to participants present in the ego file
-   **`edgelist_linked`**: one row per tie; restricted to participants present in the ego file

```{r link_datasets, results='hide'}
# join LHQ variables to ego.csv
egoData_linked <- data_ego |>
  dplyr::select(networkCanvasEgoUUID, ego_id, sessionStart, sessionFinish) |>
  dplyr::left_join(
    lhq_df |> dplyr::select(ego_id, ego_gender, ego_age, ego_lang),
    by = "ego_id"
  )

alterData_linked <- data_alter |>
  dplyr::semi_join(data_ego |> dplyr::select(networkCanvasEgoUUID),
                   by = "networkCanvasEgoUUID")

edgelist_linked <- data_edgelist |>
  dplyr::semi_join(data_ego |> dplyr::select(networkCanvasEgoUUID),
                   by = "networkCanvasEgoUUID")
```

### 1.3 Sanity Checks for Linked Datasets

Before preprocessing, it is useful to verify that the three linked datasets have the expected number of rows, unique IDs, and merge behavior. These checks help catch mismatched IDs, missing joins, or duplicate participant records early.

```{r sanity_check_linked}
# 1. egoData_linked should have one row per ego
cat("Rows in egoData_linked:", nrow(egoData_linked), "\n")
cat("Unique egos in egoData_linked:", dplyr::n_distinct(egoData_linked$networkCanvasEgoUUID), "\n\n")

stopifnot(
  nrow(egoData_linked) == dplyr::n_distinct(egoData_linked$networkCanvasEgoUUID)
)

# 2. alterData_linked should contain only egos present in egoData_linked
cat("Rows in alterData_linked:", nrow(alterData_linked), "\n")
cat("Unique egos in alterData_linked:", dplyr::n_distinct(alterData_linked$networkCanvasEgoUUID), "\n\n")

stopifnot(
  all(alterData_linked$networkCanvasEgoUUID %in% egoData_linked$networkCanvasEgoUUID)
)

# 3. edgelist_linked should contain only egos present in egoData_linked
cat("Rows in edgelist_linked:", nrow(edgelist_linked), "\n")
cat("Unique egos in edgelist_linked:", dplyr::n_distinct(edgelist_linked$networkCanvasEgoUUID), "\n\n")

stopifnot(
  all(edgelist_linked$networkCanvasEgoUUID %in% egoData_linked$networkCanvasEgoUUID)
)

# 4. check whether each ego has 15 alters (expected in this protocol)
alter_counts <- alterData_linked |>
  dplyr::count(networkCanvasEgoUUID, name = "n_alters")

print(alter_counts)

if (any(alter_counts$n_alters != 15)) {
  warning("Some egos do not have exactly 15 alters. Check protocol completion/export.")
}

# 5. check for duplicated alter UUIDs within ego
dup_alters <- alterData_linked |>
  dplyr::count(networkCanvasEgoUUID, networkCanvasUUID) |>
  dplyr::filter(n > 1)

if (nrow(dup_alters) > 0) {
  warning("Duplicated networkCanvasUUID values found within ego-level alter data.")
  print(dup_alters)
} else {
  cat("No duplicated alter UUIDs within ego.\n")
}
```

```{r sanity_check_missing_ego_alter}
# missing ego_id in ego-level file
ego_id_issues <- data_ego |>
  dplyr::filter(is.na(ego_id) | trimws(ego_id) == "") |>
  dplyr::distinct(networkCanvasEgoUUID, ego_id) |>
  dplyr::mutate(issue = "Missing ego_id")

# check alter counts
alter_count_issues <- alterData_linked |>
  dplyr::count(networkCanvasEgoUUID, name = "n_alters") |>
  dplyr::filter(n_alters != 15) |>
  dplyr::left_join(
    data_ego |>
      dplyr::select(networkCanvasEgoUUID, ego_id),
    by = "networkCanvasEgoUUID"
  ) |>
  dplyr::mutate(
    issue = dplyr::case_when(
      n_alters < 15 ~ "Fewer than 15 alters",
      n_alters > 15 ~ "More than 15 alters"
    )
  )

# flags
ego_flags <- dplyr::bind_rows(
  ego_id_issues |>
    dplyr::mutate(n_alters = NA_integer_),
  alter_count_issues |>
    dplyr::select(networkCanvasEgoUUID, ego_id, n_alters, issue)
) |>
  dplyr::arrange(networkCanvasEgoUUID)

if (nrow(ego_flags) == 0) {
  cat("[OK] All egos have valid ego_id and exactly 15 alters.\n")
} else {
  warning("Some egos failed sanity checks.")
  print(ego_flags)
}
```

## 2. Alter-Level Preprocessing

This section recodes variables from their raw Network Canvas export format into analysis-ready form. These steps must be completed before constructing any compositional or structural measures.

### 2.1 Interaction Context: Dummy Columns → Single Variable

Network Canvas exports the single-select interaction context question as five separate boolean columns (one per option). We collapse them into a single character variable for analysis.

```{r recode_context}
context_cols <- c(
  "ego_alter_interaction_context_family",
  "ego_alter_interaction_context_community",
  "ego_alter_interaction_context_school",
  "ego_alter_interaction_context_work",
  "ego_alter_interaction_context_social"
)

alterData_linked <- alterData_linked %>%
  dplyr::mutate(
    across(all_of(context_cols), ~ . == "true" | . == TRUE)
  ) %>%
  dplyr::mutate(
    ego_alter_interaction_context = dplyr::case_when(
      ego_alter_interaction_context_family       ~ "family",
      ego_alter_interaction_context_community       ~ "community",
      ego_alter_interaction_context_school          ~ "school",
      ego_alter_interaction_context_work            ~ "work",
      ego_alter_interaction_context_social          ~ "social",
      TRUE                                          ~ NA_character_
    )
  )
```

### 2.2 Language Columns: String → Logical

All `alter_language_*` and `ego_alter_language_*` columns are exported as the strings `"true"` or `"false"`. Convert to R logical for arithmetic operations downstream.

```{r recode_booleans}
lang_cols <- c(
  paste0("alter_language_",
         c("english","spanish","mandarin","cantonese","filipino",
           "vietnamese","arabic","korean","japanese","other")),
  paste0("ego_alter_language_",
         c("english","spanish","mandarin","cantonese","filipino",
           "vietnamese","arabic","korean","japanese","other"))
)

alterData_linked <- alterData_linked |>
  dplyr::mutate(across(all_of(lang_cols), ~ . == "true" | . == TRUE))
```

### 2.3 Codeswitching (CS) Frequency: Empty String → NA → Numeric

Monolingual dyads skip the codeswitching stage; Network Canvas records this as an empty string rather than `NA`. Convert to numeric so that `mean(..., na.rm = TRUE)` correctly ignores these dyads when computing compositional measures.

```{r recode_cs}
alterData_linked <- alterData_linked |>
  dplyr::mutate(
    ego_alter_cs_frequency =
      dplyr::na_if(as.character(ego_alter_cs_frequency), "") |>
      as.numeric()
  )
```

### 2.4 Ordinal Columns to Numeric

```{r ordinal_to_numeric}
alterData_linked <- alterData_linked |>
  dplyr::mutate(
    ego_alter_interaction_frequency = as.numeric(ego_alter_interaction_frequency),
    ego_alter_emotional_closeness   = as.numeric(ego_alter_emotional_closeness)
  )
```

### 2.5 Language Categorization

Before constructing compositional measures, binary language indicator columns must be combined into a single analytically meaningful category variable. We do this twice: once from the `languageUsedCategory` (what language(s) does the ego use with this alter?), and once from the `languageKnownCategory` (what language(s) does the alter know overall?).

Both functions follow the same four-level logic. Any alter who uses neither focal language falls into `"Other"`. The functions are **language-agnostic**: they detect language columns via a shared prefix (`ego_alter_language_` or `alter_language_`) and assign categories based solely on the two focal language flags supplied as arguments. Any non-focal language is treated as `"Other"`. This means the same function works for any bilingual sample by changing `lang1_col` and `lang2_col`.

| lang1 (Spanish) | lang2 (English) | Category            |
|-----------------|-----------------|---------------------|
| TRUE            | FALSE           | `"Spanish"`         |
| FALSE           | TRUE            | `"English"`         |
| TRUE            | TRUE            | `"Spanish-English"` |
| FALSE           | FALSE           | `"Other"`           |

#### 2.5.1 `languageKnownCategory` — alter's own language repertoire

*What language(s) does the alter know overall?*

Derived from `alter_language_*`. This variable defines language communities among alters.

```{r classify_alter_lang_known}
classifyLanguageKnownCategory <- function(df,
                                          lang1_col   = "alter_language_spanish",
                                          lang2_col   = "alter_language_english",
                                          lang1_label = "Spanish",
                                          lang2_label = "English",
                                          none_label  = "Other") {
  miss <- setdiff(c(lang1_col, lang2_col), names(df))
  if (length(miss) > 0) stop("Missing columns: ", paste(miss, collapse = ", "))

  to_lgl <- function(x) {
    if (is.logical(x))   return(x)
    if (is.numeric(x))   return(x == 1)
    if (is.character(x)) return(tolower(x) %in% c("true","t","1","yes","y"))
    as.logical(x)
  }

  l1 <- to_lgl(df[[lang1_col]]); l1[is.na(l1)] <- FALSE
  l2 <- to_lgl(df[[lang2_col]]); l2[is.na(l2)] <- FALSE

  dplyr::case_when(
    l1 &  l2 ~ paste0(lang1_label, "-", lang2_label),
    l1 & !l2 ~ lang1_label,
   !l1 &  l2 ~ lang2_label,
    TRUE      ~ none_label
  )
}

alterData_linked <- alterData_linked |>
  dplyr::mutate(languageKnownCategory = classifyLanguageKnownCategory(alterData_linked))

janitor::tabyl(alterData_linked$languageKnownCategory)
```

#### 2.5.2 `languageUsedCategory` — ego's language use with each alter

*(What language(s) does the ego speak with this alter?)*

Derived from `ego_alter_language_*`. This variable is the basis for compositional measures of codeswitching context (§5.5).

```{r classify_ego_alter_lang}
classifyLanguageUseCategory <- function(df,
                                        lang1_col   = "ego_alter_language_spanish",
                                        lang2_col   = "ego_alter_language_english",
                                        lang1_label = "Spanish",
                                        lang2_label = "English",
                                        none_label  = "Other") {
  miss <- setdiff(c(lang1_col, lang2_col), names(df))
  if (length(miss) > 0) stop("Missing columns: ", paste(miss, collapse = ", "))

  to_lgl <- function(x) {
    if (is.logical(x))   return(x)
    if (is.numeric(x))   return(x == 1)
    if (is.character(x)) return(tolower(x) %in% c("true","t","1","yes","y"))
    as.logical(x)
  }

  l1 <- to_lgl(df[[lang1_col]]); l1[is.na(l1)] <- FALSE
  l2 <- to_lgl(df[[lang2_col]]); l2[is.na(l2)] <- FALSE

  dplyr::case_when(
    l1 &  l2 ~ paste0(lang1_label, "-", lang2_label),
    l1 & !l2 ~ lang1_label,
   !l1 &  l2 ~ lang2_label,
    TRUE      ~ none_label
  )
}

alterData_linked <- alterData_linked |>
  dplyr::mutate(languageUsedCategory = classifyLanguageUseCategory(alterData_linked))

janitor::tabyl(alterData_linked$languageUsedCategory)
```

### 2.6 Sanity Checks for Preprocessed Variables

Verify that the key derived variables should exist and have the expected formats

```{r sanity_check_preprocessed}
required_vars <- c(
  "ego_alter_interaction_context",
  "ego_alter_interaction_frequency",
  "ego_alter_emotional_closeness",
  "ego_alter_cs_frequency",
  "languageUsedCategory",
  "languageKnownCategory"
)

missing_required_vars <- setdiff(required_vars, names(alterData_linked))

if (length(missing_required_vars) > 0) {
  stop("Missing required preprocessed variables: ",
       paste(missing_required_vars, collapse = ", "))
} else {
  cat("[OK] All required preprocessed variables are present.\n")
}

# Check allowed context values
valid_contexts <- c("family", "community", "school", "work", "social")
observed_contexts <- sort(unique(stats::na.omit(alterData_linked$ego_alter_interaction_context)))
cat("Observed interaction contexts:\n")
print(observed_contexts)

if (!all(observed_contexts %in% valid_contexts)) {
  warning("Unexpected values found in ego_alter_interaction_context.")
}

# Check numeric ranges
cat("\nRange checks:\n")
cat("ego_alter_interaction_frequency:\n")
print(range(alterData_linked$ego_alter_interaction_frequency, na.rm = TRUE))

cat("ego_alter_emotional_closeness:\n")
print(range(alterData_linked$ego_alter_emotional_closeness, na.rm = TRUE))

cat("ego_alter_cs_frequency:\n")
print(range(alterData_linked$ego_alter_cs_frequency, na.rm = TRUE))

# Check category values
valid_lang_cats <- c("Spanish", "English", "Spanish-English", "Other")

observed_used <- sort(unique(stats::na.omit(alterData_linked$languageUsedCategory)))
observed_known <- sort(unique(stats::na.omit(alterData_linked$languageKnownCategory)))

cat("\nlanguageUsedCategory:\n")
print(observed_used)

cat("\nlanguageKnownCategory:\n")
print(observed_known)

if (!all(observed_used %in% valid_lang_cats)) {
  warning("Unexpected values found in languageUsedCategory.")
}
if (!all(observed_known %in% valid_lang_cats)) {
  warning("Unexpected values found in languageKnownCategory.")
}
```

All preprocessing is complete. We now save the three preprocessed datasets (`egoData_linked`, `alterData_linked`, `edgelist_linked`) to a new `preprocessed_data/` folder as a clean checkpoint, so you can reload them in future sessions without rerunning Sections 1–2. With the data cleaned and saved, we are ready to compute language-related variables (**link**).

```{r save_preprocessed_data}
# Create output folder (does nothing if it already exists)
output_dir <- "./preprocessed_data"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Write the three linked/preprocessed datasets to CSV
readr::write_csv(
  egoData_linked,
  file.path(output_dir, "egoData_linked.csv")
)

readr::write_csv(
  alterData_linked,
  file.path(output_dir, "alterData_linked.csv")
)

readr::write_csv(
  edgelist_linked,
  file.path(output_dir, "edgelist_linked.csv")
)

cat("Saved preprocessed data to:", normalizePath(output_dir), "\n")
cat(" - egoData_linked.csv   (", nrow(egoData_linked),   "rows )\n")
cat(" - alterData_linked.csv (", nrow(alterData_linked), "rows )\n")
cat(" - edgelist_linked.csv  (", nrow(edgelist_linked),  "rows )\n")
```

## 3. Tidy Alter-Level Dataset

The tidy alter dataset (one row per ego-alter dyad) is the analytic base for all compositional measures and network visualizations. It retains only the columns needed for downstream analysis.

```{r tidy_alter}
tidy_alter <- alterData_linked |>
  dplyr::select(
    networkCanvasEgoUUID, networkCanvasUUID, nodeID,   # linking variables
    name, alter_gender, alter_age,                     # alter attributes
    ego_alter_interaction_frequency,                   # ego-alter attributes
    ego_alter_emotional_closeness,
    ego_alter_interaction_context,
    languageUsedCategory,                              # derived language categories
    languageKnownCategory,
    ego_alter_cs_frequency                             # outcome (NA = monolingual dyad)
  )

dplyr::glimpse(tidy_alter)
```

## 4. Build `egor` Object

The `egor` package provides a unified data structure for egocentric network data that keeps all three levels linked. We use `threefiles_to_egor()`, which stores alter data and alter-alter ties as nested list columns (`.alts` and `.aaties`) within a single object.

This object is required for `egor::composition()` (§5.4) and `egor::as_igraph()` (in `BiNet_network_visualization.qmd`).

```{r lang_colors}
# Color scheme for ego-alter language use categories
# Update labels to match your participant population's focal languages
lang_colors <- c(
  "Spanish"              = "#D73027",
  "English"              = "#2C5F8A",
  "Spanish-English"      = "#A8C8E8",
  "Other"                = "gray"
)
```

```{r build_egor}
alter_for_egor <- alterData_linked |>
  dplyr::mutate(
    languageKnownCategory = factor(languageKnownCategory,
                                   levels = names(lang_colors))
  )

egor_obj <- egor::threefiles_to_egor(
  egos      = egoData_linked,
  alters.df = alter_for_egor,
  edges     = edgelist_linked,
  ID.vars   = list(
    ego    = "networkCanvasEgoUUID",
    alter  = "networkCanvasUUID",
    source = "networkCanvasSourceUUID",
    target = "networkCanvasTargetUUID"
  )
)
```

## 5. Compositional Measures

Compositional measures summarize *what* an ego's network is made of by aggregating alter-level variables up to the ego level. They are constructed following a four-step pipeline:

1.  **Define a grouping variable** (an alter attribute, an ego-alter attribute, or a combination)
2.  **Compute a summary statistic** (mean or proportion) within each group for each ego
3.  **Pivot wide** so that each group becomes a separate ego-level column
4.  **Join** the result back to `egoData_linked` via `networkCanvasEgoUUID`

When no grouping variable is used (§5.1–§5.3), this reduces to a single per-ego mean, which is structurally equivalent to a global self-report item on a standard LHQ. Introducing grouping variables (§5.5 onward) produces context-specific summaries that capture within-ego variation invisible to a single global item.

### 5.1 Global CS Frequency Mean

*The simplest compositional measure: a single per-ego mean across all alters.*

When based on a single alter attribute with no grouping, this is structurally equivalent to a global CS frequency item from a standard LHQ, which summarizes the ego's average CS tendency across their whole network.

```{r comp_global_cs}
# Note: na.rm = TRUE because monolingual dyads have NA cs_frequency by design
global_cs <- tidy_alter |>
  dplyr::group_by(networkCanvasEgoUUID) |>
  dplyr::summarise(
    mean_cs_global = mean(ego_alter_cs_frequency, na.rm = TRUE),
    n_cs_dyads     = sum(!is.na(ego_alter_cs_frequency)),
    .groups = "drop"
  )

egoData_linked <- egoData_linked |>
  dplyr::left_join(global_cs, by = "networkCanvasEgoUUID")

global_cs
```

### 5.2 Global Interaction Frequency Mean

A parallel global measure can be created for interaction frequency by averaging how often the ego interacts with all nominated alters.

```{r comp_global_interaction}
global_interaction <- tidy_alter |>
  dplyr::group_by(networkCanvasEgoUUID) |>
  dplyr::summarise(
    mean_interaction_global =
      mean(ego_alter_interaction_frequency, na.rm = TRUE),
    .groups = "drop"
  )

egoData_linked <- egoData_linked |>
  dplyr::left_join(global_interaction, by = "networkCanvasEgoUUID")

global_interaction
```

### 5.3 Emotional Closeness

Another global compositional measure is emotional closeness, computed as the mean closeness rating across all alters.

```{r comp_global_closeness}
global_closeness <- tidy_alter |>
  dplyr::group_by(networkCanvasEgoUUID) |>
  dplyr::summarise(
    mean_closeness_global =
      mean(ego_alter_emotional_closeness, na.rm = TRUE),
    .groups = "drop"
  )

egoData_linked <- egoData_linked |>
  dplyr::left_join(global_closeness, by = "networkCanvasEgoUUID")

global_closeness
```

### 5.4 Alter Language Composition

*What proportion of each ego's alters belong to each language category?*

`egor::composition()` computes proportional frequencies of a categorical alter attribute for every ego in one call, returning a wide ego-level tibble. This is equivalent to the manual `group_by → summarise → pivot_wider` pipeline but more concise.

```{r comp_lang_egor}
lang_comp_ego <- egor::composition(egor_obj, alt.attr = "languageKnownCategory")
lang_comp_ego

egoData_linked <- egoData_linked |>
  dplyr::left_join(
    lang_comp_ego |> dplyr::rename(networkCanvasEgoUUID = .egoID),
    by = "networkCanvasEgoUUID"
  )
```

### 5.5 CS Frequency by Interaction Context

*Does the ego codeswitch more in one context than another?*

**Mean CS frequency grouped by the primary context of interaction.** When grouped by context, the same CS mean reveals within-ego contextual variation.

***Note on missing contexts.***\
Not all egos will have alters in every context. For example, undergraduate participants may have no alters in the "work" context. In such cases:\
- `n_context_* = 0` indicates that the ego has no alters in that context\
- `cs_context_* = NA` indicates that the mean cannot be computed due to no observations\
`0` reflects absence of ties, whereas `NA` reflects an undefined summary statistic.

```{r comp_context}
# Step 1: compute per-ego × context summaries
comp_context_long <- tidy_alter |>
  dplyr::group_by(
    networkCanvasEgoUUID,
    ego_alter_interaction_context
  ) |>
  dplyr::summarise(
    mean_cs       = mean(ego_alter_cs_frequency, na.rm = TRUE),
    n_per_context = dplyr::n(),
    .groups = "drop"
  )

# Step 2a: pivot mean CS wide
comp_context_mean <- comp_context_long |>
  tidyr::pivot_wider(
    id_cols      = networkCanvasEgoUUID,
    names_from   = ego_alter_interaction_context,
    values_from  = mean_cs,
    names_prefix = "cs_context_"
  )

# Step 2b: pivot counts wide
comp_context_n <- comp_context_long |>
  tidyr::pivot_wider(
    id_cols      = networkCanvasEgoUUID,
    names_from   = ego_alter_interaction_context,
    values_from  = n_per_context,
    names_prefix = "n_context_"
  )

# Step 2c: combine means + counts
comp_context_wide <- comp_context_mean |>
  dplyr::left_join(
    comp_context_n,
    by = "networkCanvasEgoUUID"
  )

# Step 3: join to egoData_linked
egoData_linked <- egoData_linked |>
  dplyr::left_join(
    comp_context_wide,
    by = "networkCanvasEgoUUID"
  )

comp_context_wide
```

### 5.6 Language Homophily

Language homophily captures how much of the ego’s network is associated with each focal language in actual ego–alter communication. Following Iniesta et al. (2024), each ego–alter tie is coded separately for each language: TRUE if the ego uses that language with the alter, and FALSE otherwise. An alter can count toward more than one language if the ego uses multiple languages with that person. For example, an alter with whom the ego uses both Spanish and English contributes to both Spanish homophily and English homophily.\
For each ego, language homophily is calculated as the proportion of nominated alters with whom the ego uses each focal language. Values range from 0 to 1. Higher values indicate that a larger share of the ego’s network uses that language in interaction.

```{r lang_homophily}
# For each ego, compute the proportion of alters with whom the ego uses
# each focal language. An alter can contribute to both language proportions

lang_homophily <- alterData_linked |>
  dplyr::group_by(networkCanvasEgoUUID) |>
  dplyr::summarise(
    n_alters_lang_homophily = dplyr::n(),

    prop_spanish_homophily = mean(ego_alter_language_spanish, na.rm = TRUE),
    prop_english_homophily = mean(ego_alter_language_english, na.rm = TRUE),

    n_spanish_homophily = sum(ego_alter_language_spanish, na.rm = TRUE),
    n_english_homophily = sum(ego_alter_language_english, na.rm = TRUE),

    .groups = "drop"
  )

egoData_linked <- egoData_linked |>
  dplyr::left_join(lang_homophily, by = "networkCanvasEgoUUID")

lang_homophily
```

### 5.7 Sanity Checks for Compositional Variables

Check compositional variables were successfully created and attached to egoData_linked

```{r sanity_check_compositional}
expected_comp_vars <- c(
  "mean_cs_global",
  "n_cs_dyads",
  "mean_interaction_global",
  "mean_closeness_global",
  "Spanish",
  "English",
  "Spanish-English",
  "Other",
  "cs_context_family",
  "cs_context_community",
  "cs_context_school",
  "cs_context_work",
  "cs_context_social",
  "n_alters_lang_homophily",
  "prop_spanish_homophily",
  "prop_english_homophily",
  "n_spanish_homophily",
  "n_english_homophily"
)

existing_comp_vars <- intersect(expected_comp_vars, names(egoData_linked))
missing_comp_vars  <- setdiff(expected_comp_vars, names(egoData_linked))

cat("Existing compositional variables:\n")
print(existing_comp_vars)

if (length(missing_comp_vars) > 0) {
  cat("\nMissing compositional variables (may be absent if no such alters/categories exist in this dataset):\n")
  print(missing_comp_vars)
}

egoData_linked |>
  dplyr::select(networkCanvasEgoUUID, dplyr::any_of(existing_comp_vars)) |>
  dplyr::glimpse()
```

## 6. Save Data Outputs

```{r save_outputs}
output_dir <- "./Compositional measures"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Save tidy alter-level data
readr::write_csv(
  tidy_alter,
  file.path(output_dir, "binet_tidy_alter.csv")
)

# Save the full linked ego-level dataset
readr::write_csv(
  egoData_linked,
  file.path(output_dir, "binet_egoData_linked.csv")
)

# Save a clean ego-level compositional dataset containing only variables that exist
comp_vars_to_save <- c(
  "networkCanvasEgoUUID",
  "ego_id",
  "ego_gender",
  "ego_age",
  "ego_lang",
  "mean_cs_global",
  "n_cs_dyads",
  "mean_interaction_global",
  "mean_closeness_global",
  "Spanish",
  "English",
  "Spanish-English",
  "Other",
  "cs_context_family",
  "cs_context_community",
  "cs_context_school",
  "cs_context_work",
  "cs_context_social",
  "n_alters_lang_homophily",
  "prop_spanish_homophily",
  "prop_english_homophily",
  "n_spanish_homophily",
  "n_english_homophily"
  )

ego_compositional <- egoData_linked |>
  dplyr::select(dplyr::any_of(comp_vars_to_save))

readr::write_csv(
  ego_compositional,
  file.path(output_dir, "binet_ego_compositional.csv")
)

cat("Saved to:", normalizePath(output_dir), "\n")
cat("  binet_tidy_alter.csv         —", nrow(tidy_alter), "rows\n")
cat("  binet_egoData_linked.csv     —", nrow(egoData_linked), "rows,", ncol(egoData_linked), "columns\n")
cat("  binet_ego_compositional.csv  —", nrow(ego_compositional), "rows,", ncol(ego_compositional), "columns\n\n")

cat("Columns in binet_ego_compositional.csv:\n")
cat(paste0("  ", names(ego_compositional)), sep = "\n")
```
