---
title: "Language Structural Measures Guideline"
author: "Zoey Li"
date: "`r Sys.Date()`"
format:
  gfm:
    toc: true
    toc-depth: 3
    code-overflow: wrap
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# BiNet Structural Measures Guideline

## Package dependencies and version requirements

This guideline relies on the R packages \`egor\` and \`igraph\` for network construction and computation.\
\
To ensure reproducibility and consistency of results, we recommend using the following versions:\
- egor (\>= 1.25.10)\
- igraph (\>= 2.3.0)\
\
For full reproducibility, it is strongly recommended to use the exact versions:\
- egor (= 1.25.10)\
- igraph (= 2.3.0)

Users can check their installed package versions using: packageVersion("egor") packageVersion("igraph")

```{r}
packageVersion("egor") 
packageVersion("igraph")
```

# 1. egoLangBetweenness

Compute ego language betweenness from personal network data.

This package provides tools to quantify how often an ego serves as a bridge between alters from different language communities within an ego network.

------------------------------------------------------------------------

## 1.1 Overview

In bilingualism research, language use is often embedded in social networks. Individuals interact with people who may differ in their language preferences, and these patterns shape opportunities for code-switching and language control.

**Ego language betweenness** (Tiv et al., 2022) captures the extent to which an individual (ego) connects otherwise separated language communities in their personal network.

Conceptually, this measure answers:

> *How often does the ego lie on shortest paths between alters from different language communities?*

-   A value of **0** indicates that the ego does not bridge across language communities

-   Higher values indicate that the ego is a central connector across communities

    [**Try the interactive example**](https://bic-lab-ucsd.github.io/PNS-Tutorial/BiNet_Structural_Measures/ego_betweenness_interactive_example.html) — play with the network to see how the score is computed step by step.

This measure complements existing indices such as language entropy by focusing on **network structure rather than usage proportions**.

------------------------------------------------------------------------

## 1.2 Installation

To use the package, you need to install the development version from GitHub on R:

```{r, warning=FALSE, message=FALSE}
# Install devtools package if you never installed this before
if(!"devtools" %in% rownames(installed.packages())) install.packages("devtools")

# Install the development version from github
devtools::install_github("bic-lab-ucsd/PNS-Tutorial", 
                         subdir = "BiNet_Structural_Measures/betweenness_package",
                         force = TRUE)
```

## 1.3 Usage

This package works on ego-network data. It assumes that the input data have been preprocessed, such that each alter is already assigned to a language category (e.g., `languageUsedCategory`). For example, users can refer to[preprocessing pipeline](https://github.com/bic-lab-ucsd/PNS-Tutorial/blob/main/BiNet_preprocessing_compositional_Measures/BiNet_preprocessing_tutorial.md) for a workflow

The input typically consists of:

-   `ego` → defines **who the participants are**
-   `alter` → defines **who is in each ego’s network + their attributes**
-   `edges` → defines **how alters are connected to each other**

To compute ego language betweenness, the dataset must minimally include the variables listed in the table below:

+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| Dataset   | Variable                  | Description                                            | Role in analysis                                                                                                                                   |
+===========+===========================+========================================================+====================================================================================================================================================+
| **ego**   | `networkCanvasEgoUUID`    | Unique identifier for each ego                         | Used to define and iterate over ego networks                                                                                                       |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| **ego**   | `ego_id`                  | Unique participant identifier                          | Unique identifier for each participant, used to integrate network science questionnaire data with additional questionnaires and experimental data. |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| **alter** | `networkCanvasEgoUUID`    | Ego ID for each alter                                  | Links alters to their ego                                                                                                                          |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| **alter** | `networkCanvasUUID`       | Unique identifier for each alter                       | Node identifier in the network                                                                                                                     |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| **alter** | `languageUsedCategory`    | Language community (e.g., English, Spanish, Bilingual) | Defines community membership for betweenness                                                                                                       |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| **edges** | `networkCanvasEgoUUID`    | Ego ID for each edge                                   | Ensures edges are within the same network                                                                                                          |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| **edges** | `networkCanvasSourceUUID` | Source alter ID                                        | Defines edge (node–node connection)                                                                                                                |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+
| **edges** | `networkCanvasTargetUUID` | Target alter ID                                        | Defines edge (node–node connection)                                                                                                                |
+-----------+---------------------------+--------------------------------------------------------+----------------------------------------------------------------------------------------------------------------------------------------------------+

### Example data

Below is an example using demo data included in the package. Suppose we have ten ego networks, where alters are categorized as `"English"`, `"Spanish"`, or `"Spanish-English"`.

### Step 1: load preprocessed dataset and package

```{r reload_preprocessed_data}
# Folder where the preprocessed CSVs are stored
input_dir <- "./preprocessed_data"

egoData_linked <- readr::read_csv(
  file.path(input_dir, "egoData_linked.csv"),
  show_col_types = FALSE
)

alterData_linked <- readr::read_csv(
  file.path(input_dir, "alterData_linked.csv"),
  show_col_types = FALSE
)

edgelist_linked <- readr::read_csv(
  file.path(input_dir, "edgelist_linked.csv"),
  show_col_types = FALSE
)

cat("Reloaded preprocessed data from:", normalizePath(input_dir), "\n")
cat(" - egoData_linked   (", nrow(egoData_linked),   "rows )\n")
cat(" - alterData_linked (", nrow(alterData_linked), "rows )\n")
cat(" - edgelist_linked  (", nrow(edgelist_linked),  "rows )\n")
```

Load the necessary packages. This includes the core network analysis package [**igraph, egor**]{.underline} and the **egoLangBetweenness** package.

```{r, warning=FALSE, message=FALSE}
# Install egor package if you never installed this before
if(!"egor" %in% rownames(installed.packages())) install.packages("egor")
if(!"igraph" %in% rownames(installed.packages())) install.packages("egor")

library(igraph)
library(egor)
library(egoLangBetweenness)
```

### Step 2: Compute ego language betweenness

We compute ego language betweenness for each ego network using the function\
`ego_language_betweenness_dataset()`.

This function takes three dataframes as input:

-   `ego_df`: ego-level data (participants)

-   `alter_df`: alter-level data (network members + language category)

-   `edge_df`: edge-level data (connections between alters)

```{r, warning=FALSE, message=FALSE}
betweenness_df <- ego_language_betweenness_dataset(
  ego_df = egoData_linked,
  alter_df = alterData_linked,
  edge_df = edgelist_linked
)
```

This step automatically:

-   Constructs an ego network for each participant

-   Identifies language communities based on `languageUsedCategory`

-   Computes shortest paths between alters

-   Calculates how often the ego lies on shortest paths **between alters from different language communities**

### Step 3: Inspect the output

The resulting dataframe contains one row per ego:

```{r}
print(betweenness_df)
```

The output dataframe contains one row per ego, along with variables capturing both network identifiers and key quantities involved in the computation of ego language betweenness:

+-----------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Column name**                   | **Description**                                                                                                                                                                                                                                                             |
+-----------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `networkCanvasEgoUUID`            | Unique identifier for each ego network                                                                                                                                                                                                                                      |
+-----------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `ego_id`                          | Participant ID (can be used to merge with other datasets)                                                                                                                                                                                                                   |
+-----------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `ego_language_betweenness`        | The computed betweenness value for the ego                                                                                                                                                                                                                                  |
+-----------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `n_valid_pairs`                   | Total number of alter pairs from different language communities considered in the computation                                                                                                                                                                               |
|                                   |                                                                                                                                                                                                                                                                             |
|                                   | `n_valid_pairs` can be used as a normalization factor when ego networks differ in size or in the number of cross-language alter pairs, ensuring that betweenness reflects the proportion of bridging opportunities realized rather than the absolute number of connections. |
+-----------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| `language_betweenness_normalized` | Ego language betweenness normalized by the number of cross-language alter pairs, computed as `ego_language_betweenness` / `n_valid_pairs`.                                                                                                                                  |
|                                   |                                                                                                                                                                                                                                                                             |
|                                   | This rescales the raw score into the proportion of cross-language bridging opportunities actually realized by the ego, allowing comparison across networks of different sizes or compositions.                                                                              |
|                                   |                                                                                                                                                                                                                                                                             |
|                                   | NaN indicates that the ego's network contains no alter pairs from different language communities (e.g., a fully monolingual network), making normalization undefined.                                                                                                       |
+-----------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

# 2. Language E-I Index

## 2.1 Overview

*The extent to which alter-alter ties cross language group boundaries versus stay within them.*

To quantify whether an ego’s network is organized into language-based clusters, we use the **E–I index** from social network analysis. The index compares the number of ties between alters in different language groups (**external ties**) with the number of ties between alters in the same language group (**internal ties**):

$$\text{E-I} = \frac{E - I}{E + I}$$

where $E$ is the number of ties between alters in **different** language groups and $I$ is the number of ties between alters in the **same** language group. The index ranges from **−1**.

-   **−1** = all ties are internal, indicating strong **language compartmentalization**

-   **+1** = all ties are external, indicating strong **language integration**

-   **0** = equal numbers of internal and external ties

Here, groups are defined by `languageKnownCategory` (the alter's own language repertoire). A negative E-I index means the ego's network is organized into language-homogeneous clusters; a positive value means alters regularly interact across language boundaries. When only one language group is present (e.g., a fully bilingual network), the index is `NA`.

The standard E-I index uses raw tie counts, which makes it sensitive to group size imbalance — a group with more members has more possible within-group pairs. We therefore compute a **density-weighted** version: $E$ and $I$ are replaced by between-group and within-group tie *densities* ($d_{\text{between}}$ and $d_{\text{within}}$), correcting for the number of possible ties in each stratum. Both the raw and density-weighted versions are returned.

------------------------------------------------------------------------

## 2.2 Usage

### Example data

+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| Dataset    | Variable                  | Description                                        | Role in analysis                                                             |
+============+===========================+====================================================+==============================================================================+
| ego        | `networkCanvasEgoUUID`    | Unique identifier for each ego                     | Used to define and iterate over ego networks                                 |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| ego        | `ego_id`                  | Unique participant identifier                      | Used to merge with external datasets (e.g., questionnaires, behavioral data) |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| alter      | `networkCanvasEgoUUID`    | Ego ID for each alter                              | Links alters to their corresponding ego                                      |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| alter      | `networkCanvasUUID`       | Unique identifier for each alter                   | Node identifier in the network                                               |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| alter      | `languageKnownCategory`   | Language group (e.g., English, Spanish, Bilingual) | Defines group membership for computing internal vs external ties             |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| edges      | `networkCanvasEgoUUID`    | Ego ID for each edge                               | Ensures edges are within the same ego network                                |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| edges      | `networkCanvasSourceUUID` | Source alter ID                                    | Defines edge (node–node connection)                                          |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+
| edges      | `networkCanvasTargetUUID` | Target alter ID                                    | Defines edge (node–node connection)                                          |
+------------+---------------------------+----------------------------------------------------+------------------------------------------------------------------------------+

This function works on **ego-network data**, using the similar data structure as ego language betweenness but here we focus on the `languageKnownCategory`

Below is an example using demo data included in the demo. Suppose we have ten ego networks, where alters are categorized as `"English"`, `"Spanish"`, or `"Spanish-English"`in their `languageKnownCategory`

### Step 1: load preprocessed dataset

```{r reload_preprocessedData}
# Folder where the preprocessed CSVs are stored
input_dir <- "./preprocessed_data"

egoData_linked <- readr::read_csv(
  file.path(input_dir, "egoData_linked.csv"),
  show_col_types = FALSE
)

alterData_linked <- readr::read_csv(
  file.path(input_dir, "alterData_linked.csv"),
  show_col_types = FALSE
)

edgelist_linked <- readr::read_csv(
  file.path(input_dir, "edgelist_linked.csv"),
  show_col_types = FALSE
)

cat("Reloaded preprocessed data from:", normalizePath(input_dir), "\n")
cat(" - egoData_linked   (", nrow(egoData_linked),   "rows )\n")
cat(" - alterData_linked (", nrow(alterData_linked), "rows )\n")
cat(" - edgelist_linked  (", nrow(edgelist_linked),  "rows )\n")
```

Load the necessary packages. This includes the core network analysis package [**egor**]{.underline} and the [**dplyr**]{.underline} package.

We use `egor` because `egor` provides a built-in function `EI()` that directly computes the E–I index. This avoids:

-   manually counting internal/external ties

-   manually computing densities

-   writing custom graph traversal code

In other words: `egor` handles the **network construction + EI computation internally**

```{r}
# Install egor package if you never installed this before
if(!"egor" %in% rownames(installed.packages())) install.packages("egor")
library(egor)
library(dplyr)
```

### Step 2: Build the egor object

This step converts the three data tables into a single **egor object**, storing:

-   ego attributes

-   alter attributes

-   alter–alter connections

This is necessary because the `EI()` function **requires an egor object** as input. It also ensures that each ego network is **properly separated** and standardizes the structure for network computations

```{r Build the egor object , warning=FALSE, message=FALSE}
egor_obj <- egor::threefiles_to_egor(
  egos      = egoData_linked,
  alters.df = alterData_linked,
  edges     = edgelist_linked,
  ID.vars = list(
    ego    = "networkCanvasEgoUUID",
    alter  = "networkCanvasUUID",
    source = "networkCanvasSourceUUID",
    target = "networkCanvasTargetUUID"
  )
)
```

### Step 3: Compute E-I Index

We compute two versions of the E–I index:

-   **Raw E–I index**: based on the number of ties, defined as ( (E - I) / (E + I) ) where (E) and (I) are the numbers of between-group and within-group ties. This follows the standard definition but is sensitive to group size imbalance.

-   **Density-rescaled E–I index**: replaces tie counts with tie densities (i.e., adjusts for the number of possible ties), providing a measure that is less affected by differences in group size.

Both versions are reported to capture complementary aspects of network structure.

```{r Compute E-I Index}
# Density-rescaled EI (ei_index_density)
lang_ei_index <- egor::EI(
  egor_obj,
  alt.attr    = "languageKnownCategory",
  include.ego = FALSE,
  rescale     = TRUE
) |>
  dplyr::select(.egoID, ei_index_density = ei)


ego_data <- as_tibble(egor_obj$ego)

ei_df <- lang_ei_index |>
  dplyr::rename(networkCanvasEgoUUID = .egoID) |>
  dplyr::left_join(
    ego_data |> dplyr::select(.egoID, ego_id),  
    by = c("networkCanvasEgoUUID" = ".egoID")
  )
```

```{r}
print(ei_df)
```

The output dataframe contains one row per ego, along with variables capturing both network identifiers and key quantities involved in the computation of EI index:

+------------------------+------------------------------------------------------------------------------------------------------------------------+
| Column name            | Description                                                                                                            |
+========================+========================================================================================================================+
| `networkCanvasEgoUUID` | Unique identifier for each ego network                                                                                 |
+------------------------+------------------------------------------------------------------------------------------------------------------------+
| `ego_id`               | Participant ID (can be used to merge with other datasets)                                                              |
+------------------------+------------------------------------------------------------------------------------------------------------------------+
| `lang_ei_index`        | EI index normalized by network density;                                                                                |
|                        |                                                                                                                        |
|                        | NaN indicates ego's network is fully homophilous (all alters share the same language), making normalization undefined. |
+------------------------+------------------------------------------------------------------------------------------------------------------------+

## Combine and Save the Results

```{r}
# Merge the two dataframes by networkCanvasEgoUUID
networkscience_structural_measures <- merge(betweenness_df, ei_df, 
                                            by = "networkCanvasEgoUUID")

# Save as CSV
write.csv(networkscience_structural_measures, 
          "networkscience_structural_measures.csv", 
          row.names = FALSE)
```
